How to Avoid OOTA Without Really Trying

### 2024-09-12 paper/ootavacuous.tex
ootavacuous: Compress Section 1.1 ("Brief OOTA Overview")
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

We begin with a brief overview of the OOTA problem followed by an equally
brief summary of prior work in this area.

@@

In broad terms,
OOTA occurs theoretically when a group of threads load from each others' stores
and each thread's store depends on the value returned by that thread's load.
The collection of loads and stores forms an \emph{OOTA cycle}.
In the most extreme cases a nonsensical value can pop up ``out of thin air'';
however, as shown by
Listing~\ref{lst:OOTA Cycle} below,
OOTA cycles need not involve nonsensical values.

\subsubsection{Simple OOTA Cycle}
\label{sec:Simple OOTA Cycle}

\begin{listing}[tbp]
% Constructed by hand.
{
\small
\begin{tabular}{r|l|l}
& \multicolumn{1}{c}{P0} & \multicolumn{1}{c}{P1} \\
\hline
1. &
\texttt{~~int r1 = x} &
	\texttt{~~int r2 = y} \\
2. &
\texttt{~~y = r1} &
	\texttt{~~x = r2} \\
\end{tabular}

@@

Listing~\ref{lst:Simple OOTA} shows a simple example where an OOTA
cycle might result in all of \co{x}, \co{y}, \co{r1}, and \co{r2}
having final values of 42, despite the fact that there is nothing in
the default-zero initial values or the executable code to support such
an outcome~\cite{PaulEMcKenney2020RelaxedGuideRelaxed}.
Throughout this paper, variables whose names are ``\co{r}'' followed by
a number denote registers, and other names denote shared variables.
All accesses to shared variable are relaxed atomic accesses.
First, P0's line~1 does a relaxed atomic load from \co{x} into \co{r1},
reading the value of P1's line-2 store rather than \co{x}'s initial
value of zero and somehow obtaining 42.
Second, P0's line~2 stores \co{r1}, and thus 42, to \co{y}.
Third, P1's line~1 loads 42 from \co{y} to \co{r2}.
Fourth and finally, P1's line~2 stores 42 to \co{x}, justifying the value
loaded by P0's line~1.

Because nothing else in the C++ memory model rules out such OOTA cycles
(as indicated by the ``sometimes satisfied'' in the figure), the C++
standard explicitly prohibits them in 33.5.4p8
(\co{[atomics.order]})~\cite{ThomasKoeppe2023N4950}:
\begin{quote}
	Implementations should ensure that no “out-of-thin-air” values
	are computed that circularly depend on their own computation.
\end{quote}
The standard's prohibition of OOTA is of course important, for example, to
prevent misapplication of the as-if rule in oracular C++ implementations,
but those of us writing code in the real world must rely on actual C++
implementations.
And in these implementations, this prohibition is in fact enforced by TSO
ordering in strongly ordered systems and by data-dependency ordering in
weakly ordered systems.
The need to prohibit simple OOTA is one reason why compiler-based value
speculation optimizations require checks on such speculation, and further
requires that these checks must be based on the values actually loaded.

In Listing~\ref{lst:Simple OOTA}
there is a \emph{semantic dependency} from line~1 to line~2 in both
P0 and P1.
(Roughly speaking, there is a semantic dependency from a given load
to a given store when \emph{all other things being equal, a change in the
value loaded can change the value stored or prevent the store from
occurring at all.}
Here the dependencies are trivial because the values stored simply
\emph{are} the values that were loaded.)
Since real-world CPUs cannot store something
until they have determined its value,\footnote{
	Another way of saying this is that real-world CPUs do not
	make their stores visible to other CPUs until those stores
	are no longer speculative.}
the stores on line~2 cannot take place until the P0's and P1's CPUs
know what values are returned by the loads on line~1.
Thus the hardware orders these stores after their corresponding loads,
and this ordering prevents the OOTA result.

\subsubsection{Simple Reordering}
\label{sec:Simple Reordering}

\begin{listing}[tbp]
% Constructed by hand.
{
\small
\begin{tabular}{r|l|l}
& \multicolumn{1}{c}{P0} & \multicolumn{1}{c}{P1} \\
\hline
1. &
\texttt{~~int r1 = x} &
	\texttt{~~int r2 = y} \\
2. &
\texttt{~~y = r1} &
	\texttt{~~x = 42} \\
\end{tabular}

@@

It is important to distinguish true OOTA cycles from OOTA-like
behavior caused by simple reordering, such as that shown in
Listing~\ref{lst:Simple Reordering}~\cite{PaulEMcKenney2020RelaxedGuideRelaxed}.
Both the C++ compiler and the CPU are within their rights to reorder P1's
lines~1 and~2, which can result in all of \co{x}, \co{y}, \co{r1},
and \co{r2} having the value 42 via the following steps.
First, P1's line~2 stores 42 to \co{x}.
Second, P0's line~1 loads 42 from \co{x} into \co{r1}.
Third, P0's line~2 stores \co{r1}, and thus 42, to \co{y}.
Fourth and finally, P1's line~1 loads 42 from \co{y} to \co{r2}.
Current C++ implementations can and do exhibit this reordering behavior,
as indicated by the ``sometimes satisfied'' in the figure.

\subsubsection{Non-Trivial OOTA}
\label{sec:Non-Trivial OOTA}

This paper follows P2055R0~\cite{PaulEMcKenney2020RelaxedGuideRelaxed}
in using \emph{full C++} to denote the standard including the
prohibition of OOTA mentioned above.
Unlike that article, we use \emph{loose C++}
(rather than \emph{strict C++}, which seems too similar to
\emph{full C++} for comfort)
to denote a hypothetical standard that excludes this prohibition but
is otherwise identical to full C++.
Unqualified \emph{C++} means full C++.

@@

Listing~\ref{lst:OOTA Cycle}
shows a code fragment that under loose C++ has an OOTA cycle, though
the cycle is of course prohibited in full C++.
This example differs from that of Section~\ref{sec:Simple OOTA Cycle}
in that line~3 of both processes stores the value 42, but only if
line~2 of a given process determines that the value loaded was 42,
keeping in mind that variables are default-initialized to zero.

This OOTA cycle is constructed in the following manner.
First, P0's and P1's line~1 load from \co{y} and \co{x}, respectively,
both somehow returning the value 42.
Second, P0's and P1's line~2 determine that the values loaded were both
42, so each process advances to its line~3.
Third, P0's and P1's line~3 stores the value 42 that was (somehow)
clairvoyantly loaded by the other process's line~1.

The next section looks at how prior work has refined these issues.

@@

Dependencies aren't the only mechanism that can lead to
ordering in hardware executions.
The simplest alternative is a load reading from a store in another
thread, i.e., the rfe relation.
As mentioned in Section~\ref{sec:Simple OOTA Cycle}, a CPU does
not make the value of a store instruction available for other CPUs to
load until the store commits or some time later.
And after this happens, it takes some time for the information about
the store to travel from that CPU to others, owing to the inescapable
facts that processors have nonzero size and information cannot travel
faster than the speed of light.
Since the load instruction cannot commit until its output (the value it
reads) is fully determined, it must commit after the store it reads
from.
(To be fair, this should be considered more as a \emph{result} of
ordering than as a \emph{cause} of ordering, in that
if a load commits before a store on another thread then it
cannot possibly read from that store.)

>>>>

We begin with a brief overview of the OOTA problem followed by an equally
brief summary of prior OOTA work.

@@

\begin{listing}[bp]
% Constructed by hand.
{
\small
\begin{tabular}{r|l|l}
& \multicolumn{1}{c}{P0} & \multicolumn{1}{c}{P1} \\
\hline
1. &
\texttt{~~int r1 = x} &
	\texttt{~~int r2 = y} \\
2. &
\texttt{~~y = r1} &
	\texttt{~~x = r2} \\
\end{tabular}

@@

In broad terms, OOTA occurs when a group of threads load from each others'
stores and each thread's store depends on the value returned by that
thread's load.
The collection of loads and stores forms an \emph{OOTA cycle}.
In the most extreme cases a nonsensical value can pop up ``out of thin air''
as shown in
Listing~\ref{lst:Simple OOTA}.
Here, all of \co{x}, \co{y}, \co{r1}, and \co{r2} might have final
values of 42, despite their being no instances of 42 in the default-zero
initialization or in the executable
code~\cite{PaulEMcKenney2020RelaxedGuideRelaxed}.\footnote{
	Throughout this paper, variables whose names are ``\co{r}''
	followed by a number denote registers, and other names denote
	shared variables.
	All accesses to shared variable are relaxed atomic accesses.}
First, P0's line~1 does a relaxed atomic load from \co{x} into \co{r1},
reading the value of P1's line-2 store rather than \co{x}'s initial
value of zero and somehow obtaining 42.
Second, P0's line~2 stores \co{r1}, and thus 42, to \co{y}.
Third, P1's line~1 loads 42 from \co{y} to \co{r2}.
Finally, P1's line~2 stores 42 to \co{x}, justifying the value loaded
by P0's line~1.

Because nothing else in the C++ memory model rules out such OOTA cycles
(as indicated by the ``sometimes satisfied'' in the figure), the C++
standard explicitly prohibits them in 33.5.4p8
(\co{[atomics.order]})~\cite{ThomasKoeppe2023N4950},
which prevents misapplication of the as-if rule in oracular C++
implementations:
\begin{quote}
	Implementations should ensure that no “out-of-thin-air” values
	are computed that circularly depend on their own computation.
\end{quote}
This paper uses \emph{full C++} to denote the standard including this
prohibition of OOTA and \emph{loose C++} to denote a hypothetical standard
that excludes this prohibition but is otherwise identical to full C++.
Unqualified \emph{C++} means full C++.

In Listing~\ref{lst:Simple OOTA}
there is a \emph{semantic dependency} from line~1 to line~2 in both
P0 and P1.
(Roughly speaking, there is a semantic dependency from a given load
to a given store when \emph{all other things being equal, a change in the
value loaded can change the value stored or prevent the store from
occurring at all.}
Here the dependencies are trivial because the values stored simply
\emph{are} the values that were loaded.)
Since real-world CPUs cannot store something
until they have determined its value,\footnote{
	Another way of saying this is that real-world CPUs do not
	make their stores visible to other CPUs until those stores
	are no longer speculative.
	This same restriction applies to compiler-based speculation.}
the stores on line~2 cannot take place until the P0's and P1's CPUs
know what values are returned by the loads on line~1.
Thus the hardware orders these stores after their corresponding loads,
and this ordering prevents the OOTA result.

\begin{listing}[bp]
% Constructed by hand.
{
\small
\begin{tabular}{r|l|l}
& \multicolumn{1}{c}{P0} & \multicolumn{1}{c}{P1} \\
\hline
1. &
\texttt{~~int r1 = x} &
	\texttt{~~int r2 = y} \\
2. &
\texttt{~~y = r1} &
	\texttt{~~x = 42} \\
\end{tabular}

@@

Simple reordering can cause non-OOTA, but OOTA-like, behavior, as
exemplified by
Listing~\ref{lst:Simple Reordering}~\cite{PaulEMcKenney2020RelaxedGuideRelaxed}.
Both the compiler and the CPU might reorder P1's lines~1 and~2, so that
all of \co{x}, \co{y}, \co{r1}, and \co{r2} might have the value 42,
as indicated by the ``sometimes satisfied'' in the figure.

@@

On the other hand, OOTA cycles can arise even when value 42 is mentioned,
as shown in
Listing~\ref{lst:OOTA Cycle}.
Line~3 of both processes stores the value 42 only when line~2 determines
that the value loaded was 42.
Default-zero initialization prevents this in both full C++ and (as we will
see) real-world loose C++, but not an oracular loose C++ implementation.

The next section looks at prior work on these issues.

@@

Dependencies aren't the only mechanism that can lead to
ordering in hardware executions.
The simplest alternative is a load reading from a store in another
thread, i.e., the rfe relation.
As mentioned in Section~\ref{sec:Brief OOTA Overview}, a CPU does
not make the value of a store instruction available for other CPUs to
load until the store commits or some time later.
And after this happens, it takes some time for the information about
the store to travel from that CPU to others, owing to the inescapable
facts that processors have nonzero size and information cannot travel
faster than the speed of light.
Since the load instruction cannot commit until its output (the value it
reads) is fully determined, it must commit after the store it reads
from.
(To be fair, this should be considered more as a \emph{result} of
ordering than as a \emph{cause} of ordering, in that
if a load commits before a store on another thread then it
cannot possibly read from that store.)
